### 2020-10-27 latex/vlp.tex
Major edits to the VLP chapter
The proof was lacking some detail, so I added it according to manfredi's
suggestions. It is now in a slightly better format.

Let us define the VNHC \(l = L(q,p)\), by which we mean we are
actually defining the VNHC \(h(\mathbf{q},\mathbf{p}) = l - L(q,p) = 0\) of
order 1.
The VLP satisfies
\(\nabla_q \Minv(\mathbf{q}) = \Zmat{2 \times 2}\).
By Theorem \ref{thm:vnhc-regularity},
\(h(\mathbf{q},\mathbf{p})\) is a regular VNHC whenever 
\(L(q,p)\) is \(C^2\) because
\[
   dh_{\mathbf{q}} \Minv(\mathbf{q})B = 
   \begin{bmatrix}
      -\pdiff{L}{q} & 1
   \end{bmatrix}
   \begin{bmatrix}
      \frac{1}{ml^2}  & 0 \\
      0 & \frac{1}{m}
   \end{bmatrix} 
   \begin{bmatrix}
      0 \\ 
      1
   \end{bmatrix}
   = \frac{1}{m}
\]
is always full rank.
The constraint manifold is
\(\Gamma = \mathbb{S}^1 \times \R\), which is parameterized by the unactuated
phase \((q,p)\).
The constrained dynamics are Hamiltonian, and Theorem \ref{thm:zero-dynamics}
tells us they are described entirely by \((\dot{q},\dot{p})\) with \(l\)
replaced by \(L(q,p)\):
\begin{align}\label{eqn:vlp-hamiltonian}
   \mathcal{H}(q,p) &= \frac{1}{2} \frac{p^2}{mL^2} - \frac{1}{2}\dot{L}^2 -
      mgL\cos(q)
   , \\
  &\begin{cases}
      \dot{q} = \frac{p}{m L^2} \\
      \dot{p} = -mgL\sin(q)
      . \\ 
   \end{cases} \nonumber
\end{align}
Note that we suppress the function notation of \(L(q,p)\) and
\(\dot{L}(q,p)\) for clarity.

@@

\begin{defn}\label{defn:energy-injection}
   Let \(D \subset \R^2\) with \(0 \in D\).
   Let \(f : D \rightarrow \R^2\) be locally Lipschitz with \(f(0) = 0\).
   The system described by \(\dot{x} = f(x)\) 
   \textit{gains energy on \(D\)} if 
   \begin{enumerate}
      \item The origin is a repeller 
         (or equivalently, the origin is asymptotically stable in negative time).
      \item For all compact sets \(K \subset D\) and for almost every initial
         condition \(x(0) \in K\), there exists \(T > 0\) such
         that \(x(t) \not \in K \, (\forall t > T)\).
   \end{enumerate}
   The system \textit{loses energy on \(D\)} if it gains energy in
   negative-time.
\end{defn}

@@

   Choose, as a candidate Lyapunov function, the energy for the
   average-length pendulum with zero-potential at the bottom of the swing:
   \[
       E_\text{avg}(q,p) := \frac{1}{2}\frac{p^2}{m l_\text{avg}^2} 
           + m g l_\text{avg} (1-\cos(q))
      .
   \]
   This is positive definite at \((0,0)\) and has compact sublevel
   sets on \(\Gamma\). 
   The derivative of \(E_\text{avg}\) along (\ref{eqn:negative-time-vlp}) is
   \[
      \dot{E}_\text{avg} = \frac{g\sin(q)p \left(L(\theta)^3 - l_\text{avg}^3\right)}
              {l_\text{avg}^2L(\theta)^2}
      .
   \]
   The sign of \(\sin(q)p\) depends only on the quadrant in the
   \((q,p)\)-plane. In fact, it is easy to show that
   \[
      \sign{\sin(q)p} = \sign{\sin(2\theta)}
      .
   \]
   Furthermore, by Lemma \ref{lemma:sign-of-cube} we have
   \[ 
       \sign{L(\theta)^3 - l_\text{avg}^3} = \sign{L(\theta) - l_\text{avg}}
       .
   \]
   The derivative of \(E_\text{avg}\) is never positive since
   \begin{align}\label{eqn:vlp-proof-neg-inv}
    \sign{\dot{E}_\text{avg}} &= \sign{\sin(q)p \left(L(\theta)^3 - l_\text{avg}^3\right)}
          \nonumber \\
          &= \sign{\sin(2\theta) \left(L(\theta) - l_\text{avg}\right)}
          \nonumber \\
          &= -\sign{(l_\text{avg} - L(\theta))\sin(2\theta)} 
          \nonumber \\
          &\leq 0 \text{ (by assumption)}
      .
   \end{align}
   Thus, \(E_\text{avg}\) is a Lyapunov function, which means the negative-time system is
   stable \cite{lyapunov}.

   The linearization of (\ref{eqn:negative-time-vlp}) at the upright equilibrium
   \((\pi, 0)\) is given by
   \begin{equation}\label{eqn:linearized-vlp-neg-time}
      df_{(\pi,0)} = \begin{bmatrix}
         0 & -\frac{1}{mL(0)^2} \\
         -m g L(0) & 0 \\
      \end{bmatrix}
      ,
   \end{equation}
   which has eigenvalues at \(\lambda = \pm \sqrt{\frac{g}{L(0)}}\).
   The positive-time system (\ref{eqn:vlp-hamiltonian}) has the same eigenvalues
   at this equilibrium.
   Label the set of initial conditions converging to \((\pi, 0)\) in
   positive time by
   \[
      \Pi_0^+ := \left\{ (q_0,p_0) \in \Gamma \mid \lim_{t \to \infty} (q(t),p(t)) = (\pi, 0) \right\}
      ,
   \]
   and the set of initial conditions converging to \((\pi,0)\)
   in negative time by \(\Pi_0^-\).
   Taking the sublevel set
   \[
      E_{(\pi,0)} := \left\{ (q,p) \in \Gamma \mid E_\text{avg}(q,p) <
         E_\text{avg}(\pi,0) \right\}
      ,
   \] 
   it is clear form (\ref{eqn:vlp-proof-neg-inv}) that 
   \(\Pi_0^+ \subset E_{(\pi,0)}\)
   and 
   \(\Pi_0^- \subset \left(\Gamma \backslash E_{(\pi,0)} \right) \cup (\pi,0)\).
   Since only one eigenvalue of (\ref{eqn:linearized-vlp-neg-time})
   is negative, the stable manifold theorem tells us that both \(\Pi_0^+\) and
   \(\Pi_0^-\) are one-dimensional \cite{stable_manifold}.
   This implies the origin cannot be globally asymptotically stable for
   (\ref{eqn:negative-time-vlp}); however, we can still show it is asymptotically stable
   for almost every initial condition because \(\Pi_0^+\) and \(\Pi_0^-\) are of
   measure zero within \(\Gamma\).

   Define \(D := \Gamma \backslash \Pi_0^-\) to be the domain of interest.
   Using a Krasovskii-LaSalle argument \cite{krasovskii_lasalle}, 
   we will prove that (\ref{eqn:negative-time-vlp}) asymptotically converges to
   \((0,0)\) if it is initialized on \(D\).
   First, we define the zero-set
   \[
      Z = \left\{(q,p) \in \Gamma \mid 
      \dot{E}_\text{avg}(q,p) = 0 \right\}
      .
   \]
   It is easy to show that \(Z\) is the union of the coordinate axes
   \[
      Z= \mathbb{S}^1 \times \left\{0\right\} \cup 
         \left\{0\right\} \times \R
      .
   \]
   Krasovskii-LaSalle tells us that solutions to (\ref{eqn:negative-time-vlp})
   converge to the largest invariant set in \(Z \cap D\).
   Since \(f(q,p) = (0,0)\) on \(D\) only when \((q,p) = (0,0)\),
   the origin is asymptotically stable on \(D\) for the negative-time system.
   In particular, the origin is asymptotically stable for every initial
   condition in the interior of \(E_{(\pi,0)}\).
   Hence, the positive-time system is diverging away from the origin 
   in the \((q,p)\) plane, which means the origin is a repeller.

   We now prove the system escapes compact sets in finite time.
   Let \(x(t) = (q(t),p(t))\) be the orbit for the positive-time system.
   Let \(K \subset \Gamma\) be compact and assume \(x(0) \in K\backslash \Pi_0^+\).
   For a proof by contradiction, suppose that for all \(T > 0\) there exists
   \(t' > T\) where \(x(t') \in K\).
   Let 
   \[
      k := \max\limits_{(q,p) \in K} E_\text{avg}(q,p)
      ,
   \] 
   be the largest energy value attained in \(K\), and let \(E_k\) be its
   sublevel set.
   All sublevel sets of \(E_\text{avg}\) are compact, negatively
   invariant (by (\ref{eqn:vlp-proof-neg-inv})), and \(E_k\) contains \(K\).
   We assumed \(x(t') \in K\), so it follows that \(x(t') \in E_k\). 
   Then, since \(E_k\) is negatively invariant, the half orbit
   \(x([0,t'])\) is contained fully within \(E_k\).
   In particular, \(x(T) \in E_k\);
   but \(T\) was chosen arbitrarily, which implies \(x(t) \in E_k\) for all
   \(t \in [0,\infty[\).
   Since \(E_k\) is compact, \(x(t)\) has a positive limit set in
   \(E_k\) which (by Poincar\'{e}-Bendixson) is either an equilibrium or a
   closed orbit \cite{poincare_bendixson}.
   The system cannot have closed orbits, since this would violate asymptotic
   stability of \((0,0)\) in negative time. 
   Hence, the positive limit set must be one of the equilibria. 
   The origin is a repeller, which means \(x(t)\) must be converging to 
   \((\pi,0)\).
   This is a contradiction, since we assumed \(x(0) \not \in \Pi_0^+\).

   We have shown that orbits of the constrained dynamics will leave compact sets in
   finite time, including all sublevel sets of \(E_\text{avg}\). 
   The orbit \((q(t),p(t))\) is spiraling away from the origin and, for almost
   all initial conditions, will not converge to the upper equilibrium.
   Thus, the VLP is gaining energy on average because its momentum is increasing.
   Flipping the inequality of (\ref{eqn:vlp-energy-gain-condition}) 
   means \(E_\text{avg}\) is a Lyapunov function for the positive time system;
   by the same arguments presented here, the VLP loses energy while
   converging asymptotically to \((0,0)\).
 \end{proof}

>>>>

Let us define the VNHC \(l = L(q,p)\), by which we mean we are
actually defining the VNHC \(h(\mathbf{q},\mathbf{p}) = l - L(q,p) = 0\) of
order 1.
The VLP satisfies
\(\nabla_q \Minv(\mathbf{q}) = \Zmat{2 \times 2}\).
By Theorem \ref{thm:vnhc-regularity},
\(h(\mathbf{q},\mathbf{p})\) is a regular VNHC whenever 
\(L(q,p)\) is \(C^2\) because
\[
   dh_{\mathbf{q}} \Minv(\mathbf{q})B = 
   \begin{bmatrix}
      -\pdiff{L}{q} & 1
   \end{bmatrix}
   \begin{bmatrix}
      \frac{1}{ml^2}  & 0 \\
      0 & \frac{1}{m}
   \end{bmatrix} 
   \begin{bmatrix}
      0 \\ 
      1
   \end{bmatrix}
   = \frac{1}{m}
\]
is always full rank.
The constraint manifold is
\(\Gamma = \mathbb{S}^1 \times \R\), which is parameterized by the unactuated
phase \((q,p)\).
By Theorem \ref{thm:zero-dynamics}, the constrained dynamics
are described entirely by \((\dot{q},\dot{p})\) with \(l\)
replaced by \(L(q,p)\):
\begin{equation}\label{eqn:vlp-hamiltonian}
   \begin{cases}
      \dot{q} = \frac{p}{m L^2} \\
      \dot{p} = -mgL\sin(q)
      . \\ 
   \end{cases}
\end{equation}
Note that we suppress the function notation of \(L(q,p)\) for clarity.

@@

\begin{defn}\label{defn:energy-injection}
   Let \(D \subset \R^2\) be open with \(0 \in D\).
   Let \(f : D \rightarrow \R^2\) be locally Lipschitz with \(f(0) = 0\).
   The system described by \(\dot{x} = f(x)\) 
   \textit{gains energy on \(D\)} if 
   \begin{enumerate}
      \item The origin is a repeller 
         (or equivalently, the origin is asymptotically stable in negative time).
      \item For all compact sets \(K \subset D\) and for almost every initial
         condition \(x(0) \in K\), there exists \(T > 0\) such
         that \(x(t) \not \in K \, (\forall t > T)\).
   \end{enumerate}
   The system \textit{loses energy on \(D\)} if it gains energy in
   negative-time.
\end{defn}

@@

   Choose, as a candidate Lyapunov function, the energy for the
   average-length pendulum with zero-potential at the bottom of the swing:
   \[
       E_\text{avg}(q,p) := \frac{1}{2}\frac{p^2}{m l_\text{avg}^2} 
           + m g l_\text{avg} (1-\cos(q))
      .
   \]
   This is positive definite at \((0,0)\) and has compact sublevel
   sets on \(\Gamma\). 
   The derivative of \(E_\text{avg}\) along (\ref{eqn:vlp-hamiltonian}) is
   \begin{equation}\label{eqn:eavg-dot-pos}
      \dot{E}_\text{avg}^+ = -\frac{g\sin(q)p \left(L(\theta)^3 - l_\text{avg}^3\right)}
              {l_\text{avg}^2L(\theta)^2}
      ,
   \end{equation}
   and the derivative of \(E_\text{avg}\) along (\ref{eqn:negative-time-vlp}) is
   \begin{equation}\label{eqn:eavg-dot-neg}
      \dot{E}_\text{avg}^- = - \dot{E}_\text{avg}^+
      .
   \end{equation}
   The sign of \(\sin(q)p\) depends only on the quadrant in the
   \((q,p)\)-plane. In fact, it is easy to show that
   \[
      \sign{\sin(q)p} = \sign{\sin(2\theta)}
      .
   \]
   Furthermore, by Lemma \ref{lemma:sign-of-cube} we have
   \[ 
       \sign{L(\theta)^3 - l_\text{avg}^3} = \sign{L(\theta) - l_\text{avg}}
       .
   \]
   The positive-time derivative (\ref{eqn:eavg-dot-pos}) is nonnegative since
   \begin{align}\label{eqn:vlp-proof-neg-inv}
    \sign{\dot{E}_\text{avg}^+} &= \sign{-\sin(q)p \left(L(\theta)^3 - l_\text{avg}^3\right)}
          \nonumber \\
          &= \sign{-\sin(2\theta) \left(L(\theta) - l_\text{avg}\right)}
          \nonumber \\
          &= \sign{(l_\text{avg} - L(\theta))\sin(2\theta)} 
          \nonumber \\
          &\geq 0 \text{ (by assumption)}
      .
   \end{align}
   Likewise, the negative-time derivative (\ref{eqn:eavg-dot-neg}) is never
   positive, so \(E_\text{avg}\) is a Lyapunov function and the negative-time system is
   stable \cite{lyapunov}.

   Let us briefly analyze the upright equilibrium.
   Label the set of initial conditions converging to \((\pi, 0)\) in
   positive time by
   \[
      \Pi_0^+ := \left\{ (q_0,p_0) \in \Gamma \mid \lim_{t \to \infty} (q(t),p(t)) = (\pi, 0) \right\}
      ,
   \]
   and the set of initial conditions converging to \((\pi,0)\)
   in negative time by \(\Pi_0^-\).
   Take now the sublevel set
   \[
      E_{(\pi,0)} := \left\{ (q,p) \in \Gamma \mid E_\text{avg}(q,p) \leq
         E_\text{avg}(\pi,0) \right\}
      .
   \] 
   From (\ref{eqn:vlp-proof-neg-inv}), all sublevel sets of \(E_\text{avg}\) are
   negatively invariant in positive-time. 
   Hence, the complement \(\left(E_{(\pi,0)}\right)^C = \Gamma \backslash E_{(\pi,0)}\) is
   positively invariant.
   If the system is initialized on \(\left(E_{(\pi,0)}\right)^C\), it cannot
   posibly converge to \((\pi,0)\);
   hence, \(\Pi_0^+ \subset E_{(\pi,0)}\).
   Similarly, positive invariance of \(E_{(\pi,0)}\) in negative-time allows us
   to show that \(\Pi_0^- \subset \left(E_{(\pi,0)}\right)^C \cup (\pi,0)\).

   The linearization of (\ref{eqn:vlp-hamiltonian}) at \((\pi, 0)\) is given by
   \begin{equation}\label{eqn:linearized-vlp}
      \begin{bmatrix}
         0 & \frac{1}{mL(0)^2} \\
         m g L(0) & 0 \\
      \end{bmatrix}
      ,
   \end{equation}
   which has eigenvalues at \(\lambda = \pm \sqrt{\frac{g}{L(0)}}\).
   Linearizing (\ref{eqn:negative-time-vlp}) at this point yields the same eigenvalues.
   Since one eigenvalue of (\ref{eqn:linearized-vlp})
   is positive and one is negative, the stable manifold theorem tells
   us that both \(\Pi_0^+\) and \(\Pi_0^-\) are one-dimensional
   \cite{stable_manifold}.
   This implies the origin cannot be globally asymptotically stable for
   (\ref{eqn:negative-time-vlp}); however, we can still show it is asymptotically stable
   for almost every initial condition because \(\Pi_0^+\) and \(\Pi_0^-\) are of
   measure zero within \(\Gamma\).

   We cannot linearize the dynamics at the origin to prove it is a repeller,
   because \(\theta\) is not defined there.
   Instead, let \(D := \Gamma \backslash \Pi_0^-\).
   We will prove that (\ref{eqn:negative-time-vlp}) asymptotically converges to
   \((0,0)\) if it is initialized on \(D\).
   First, define the zero-set
   \[
      Z = \left\{(q,p) \in \Gamma \mid 
      \dot{E}_\text{avg}(q,p) = 0 \right\}
      .
   \]
   It is easy to show that \(Z\) is the union of the coordinate axes
   \[
      Z= \mathbb{S}^1 \times \left\{0\right\} \cup 
         \left\{0\right\} \times \R
      .
   \]
   Krasovskii-LaSalle tells us that solutions to (\ref{eqn:negative-time-vlp})
   converge to the largest invariant set in \(Z \cap D\)
   \cite{krasovskii_lasalle}.
   Since \(f(q,p) = (0,0)\) on \(D\) only when \((q,p) = (0,0)\),
   the origin is asymptotically stable on \(D\) for the negative-time system.
   In particular, the origin is asymptotically stable for every initial
   condition in the interior of \(E_{(\pi,0)}\).
   Hence, the positive-time system is diverging away from the origin 
   in the \((q,p)\) plane, which means the origin is a repeller.

   We now prove the system escapes compact sets in finite time.
   Let \(x(t) = (q(t),p(t))\) be the orbit for the positive-time system.
   Let \(K \subset \Gamma\) be compact and assume 
   \(x(0) \in K\backslash \Pi_0^+\) and \(x(0) \neq (0,0)\).
   For a proof by contradiction, suppose that for all \(T > 0\) there exists
   \(t' > T\) where \(x(t') \in K\).
   Let 
   \[
      k := \max\limits_{(q,p) \in K} E_\text{avg}(q,p)
      ,
   \] 
   be the largest energy value attained in \(K\), and let \(E_k\) be its
   sublevel set.
   All sublevel sets of \(E_\text{avg}\) are compact, negatively
   invariant, and \(E_k\) contains \(K\).
   We assumed \(x(t') \in K\), so it follows that \(x(t') \in E_k\). 
   Then, since \(E_k\) is negatively invariant, the half orbit
   \(x([0,t'])\) is contained fully within \(E_k\).
   In particular, \(x(T) \in E_k\);
   but \(T\) was chosen arbitrarily, which implies \(x(t) \in E_k\) for all
   \(t \in [0,\infty[\).
   Since \(E_k\) is compact, \(x(t)\) has a positive limit set in
   \(E_k\) which, by Poincar\'{e}-Bendixson, is either an equilibrium or a
   closed orbit \cite{poincare_bendixson}.

   Suppose the system has a closed orbit. It would necessarily be contained in a
   level set of \(E_\text{avg}\) (since this is a non-decreasing function and
   closed orbits are periodic).
   Thus, \(E_\text{avg}\) is constant along the orbit, which means the
   orbit is contained in one of the axes because that is the only region where
   \(\dot{E}_\text{avg}^+ = 0\).
   Since the only invariant subsets of the axes are equilibria, the system cannot
   contain closed orbits.

   Hence, the positive limit set of \(x(t)\) must be one of the equilibria. 
   The origin is a repeller and \(x(0) \neq (0,0)\), which means \(x(t)\) is
   converging to \((\pi,0)\).
   This is a contradiction, since we assumed \(x(0) \not \in \Pi_0^+\).

   We conclude that the VNHC is injecting energy into the VLP.
   By flipping the inequality of (\ref{eqn:vlp-energy-gain-condition})
   we find the VLP is gaining energy in negative-time, so the VNHC would be
   dissipating energy.
\end{proof}
